A Plasma Torus Around a Young Low-Mass Star

### 2025-04-07 paper/ms.tex
update

\begin{abstract}
\normalfont
Roughly one percent of red dwarfs younger than 100 million years show
structured, periodic optical light curves suggestive of transiting
opaque material that corotates with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  However, the
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces \cite{Koen2021}.
Here, we present time-series spectroscopy and photometry of a
rapidly-rotating ($P$=3.9\,hr) CPV, TIC~141146667. The spectra show
sinusoidal time-varying Balmer emission at twice to four times the
star's equatorial velocity, directly demonstrating for the first
time that corotating clumps of circumstellar plasma exist around a CPV.
Given that cool ($10^4$ K) plasma can persist in the hot ($10^6$ K)
coronae of rapidly-rotating stars with a wide range of masses
\cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that the magnetospheres of red dwarfs can
support long-lived, optically thick clumps of material that produce
the sharp photometric features seen in CPVs.  The origin of this
optically thick material and its microphysical opacity remain
unclear.
\end{abstract}

@@

Spectra of magnetically-active, rapidly rotating stars with a wide
range of masses have been previously observed to exhibit both
sinusoidal time-varying Balmer emission
\cite{Donati2000,Townsend2005,Dunstone2006,Skelly2008} and transient
absorption features
\cite{CollierCameron1989,CollierCameron1992,Cang2020}, similar to
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \cite{Bouma2024}.   One interpretation for the
spectroscopic variability comes from an analogy to quiescent solar
prominences, which are cool condensations of plasma in the solar
corona that can last days to weeks \cite{VialEngvold2015}.  These
condensations fall back to the Sun's photosphere because gravity is
stronger than any magnetic tension or centrifugal force capable of
sustaining them.  However for stars with magnetospheric radii $R_{\rm
m}$ that exceed their corotation radii $R_{\rm c}$, the effective
potential experienced by a charged particle can have a local minimum
outside $R_{\rm c}$, enabling the material to accumulate in a
centrifugally-supported magnetosphere
\cite{Petit2013,Daley-Yates2024}.  Generally speaking, such material
need neither transit nor be opaque in broadband optical light.  In
fact, Figure~\ref{fig:spec} suggests that the transits of the
H$\alpha$ emitting material cannot explain the entirety of the complex
photometric modulation;  the spectroscopic transits happen too fast,
and at orbital phases that only partially overlap with the photometric
modulations.

Nonetheless, our Keck/HIRES observations are the first reported
time-series spectra of a CPV, and they show for the first time that
corotating circumstellar plasma clumps exist around at least one such
star.  While starspots likely do contribute smooth signals to the
photometric variability, a ``starspot-only'' scenario for the CPVs
\cite{Koen2021} seems to be incompatible with our observations.
Scenarios in which the circumstellar material is made only of dust are
similarly ruled out.  While starspots and dust may both be present,
the observed H$\alpha$ emission requires clumps of plasma with a
significant population of hydrogen atoms in the $n$=3 excited state,
with most H$\alpha$ emission in each clump coming from a region half
the size of the star (see \S~\ref{subsec:halpha}).  The observed
H$\alpha$ luminosity suggests characteristic densities and masses for
the gaseous component of these clumps of $n_{\rm H} \sim
10^{11}$\,cm$^{-3}$ and $M_{\rm gas} \sim 10^{17}$\,g (see
\S~\ref{subsec:gas}).  Dust, if present, is independently constrained
to have a total mass $10^{15}\,{\rm g} < M_{\rm dust} < 10^{17}\,{\rm
g}$ (see \S~\ref{subsec:dust}).  This circumstellar material could
originate either from the star or from an external source.  Plausible
external sources include an undetected old disk, comets, or a close-in
exoplanet.  This latter scenario would make CPVs extrasolar analogs of
the Jupiter-Io plasma torus (e.g.~Ref.~\cite{Bagenal1981}).

The other potential analog for the CPVs are the $\sigma$~Ori~E
variables, a rare subset of B stars with radiatively-driven winds
which accumulate into warped plasma tori
\cite{Townsend2005,Townsend2008}.  These tori tend to have dense
antipodal accumulations of plasma sculpted by tilted-dipole magnetic
fields, and the transits of these clumps produce broadband photometric
variability through bound-free scattering \cite{Townsend2005} and
Thomson scattering \cite{Berry2022}.  The plasma accumulates at
antipodes 180$^\circ$ apart because the deepest local minima in the
effective potential exist along the line of intersection between the
rotational and magnetic equators (see Equation~22 of
\cite{Townsend2005}).  For $\sigma$~Ori~E and almost all of its
analogs, the result is ``simple'' light curves that resemble those of
eclipsing binaries, and time-dynamic H$\alpha$ spectra similar to
those in Figure~\ref{fig:spec} \cite{Townsend2005,Townsend2008}.  The
two known exceptions, HD~37776 and HD~64740, show complex light curves
resembling CPVs \cite{Mikulasek2020,Bouma2024} and have
spectropolarimetric magnetic field maps indicating strong
contributions from higher-order magnetic moments
\cite{Kochukhov2011,Shultz2018}.  There are two implications: first,
the photometric complexity of CPVs may be a direct consequence of
magnetic fields with highly multipolar contributions.  Second, despite
this complexity, the observation from Figure~\ref{fig:spec} of two
emission clumps separated 180$^\circ$ in phase is consistent with the
expected emission from a warped plasma torus.

>>>>

\begin{abstract}
\normalfont
A small fraction of red dwarfs younger than 100 million years show
structured, periodic optical light curves suggestive of transiting
opaque material that corotates with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  However, the
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
bright or dark regions on the stellar surfaces \cite{Koen2021}.
Here, we present time-series spectroscopy and photometry of a
rapidly-rotating ($P$=3.9\,hr) CPV, TIC~141146667. The spectra show
sinusoidal time-varying Balmer emission at twice to four times the
star's equatorial velocity, which provides the first direct evidence for
corotating plasma clumps around a CPV.  While cool ($10^4$ K)
plasma can persist within the hot ($10^6$ K) coronae of
rapidly-rotating stars across a broad range of masses
\cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
our observations suggest that red dwarfs in particular can
sustain long-lived, optically thick clumps that generate the 
sharp photometric features seen in CPVs.  The origin of this
optically thick material and its microphysical opacity remain
unclear.
\end{abstract}

@@

Spectra of magnetically-active, rapidly rotating stars with a wide
range of masses have been previously observed to exhibit both
sinusoidal time-varying Balmer emission
\cite{Donati2000,Townsend2005,Dunstone2006,Skelly2008} and transient
absorption features
\cite{CollierCameron1989,CollierCameron1992,Cang2020}, similar to
Figure~\ref{fig:spec}.  No such stars were previously known to show
complex light curves \cite{Bouma2024}.   One interpretation for this
spectroscopic variability comes from an analogy to quiescent solar
prominences, which are cool condensations of plasma in the solar
corona that can last days to weeks \cite{VialEngvold2015}.  These
condensations fall back to the Sun's photosphere because gravity is
stronger than any magnetic tension or centrifugal force capable of
sustaining them.  However for stars with magnetospheric radii $R_{\rm
m}$ that exceed their corotation radii $R_{\rm c}$, the effective
potential experienced by a charged particle can have a local minimum
outside $R_{\rm c}$, enabling the material to accumulate in a
centrifugally-supported magnetosphere
\cite{Petit2013,Daley-Yates2024}.  Generally speaking, such material
need neither transit nor be opaque in broadband optical light.  In
fact, Figure~\ref{fig:spec} suggests that the transits of the
H$\alpha$ emitting material cannot explain the entirety of the complex
photometric modulation;  the spectroscopic transits happen too fast,
and at orbital phases that only partially overlap with the photometric
modulations.

Nonetheless, our Keck/HIRES observations are the first reported
time-series spectra of a CPV, and they demonstrate that
corotating circumstellar plasma clumps exist around at least one such
star.  While starspots likely do contribute smooth signals to the
photometric variability, a ``starspot-only'' scenario for the CPVs
\cite{Koen2021} is incompatible with our observations.
Scenarios in which the circumstellar material is made only of dust are
similarly ruled out.  While starspots and dust may both be present,
the observed H$\alpha$ emission requires clumps of plasma with a
significant population of hydrogen atoms in the $n$=3 excited state,
with most H$\alpha$ emission in each clump coming from a region half
the size of the star (see \S~\ref{subsec:halpha}).  The observed
H$\alpha$ luminosity suggests characteristic densities and masses for
the gaseous component of these clumps of $n_{\rm H} \sim
10^{11}$\,cm$^{-3}$ and $M_{\rm gas} \sim 10^{17}$\,g (see
\S~\ref{subsec:gas}).  Dust, if present, is independently constrained
to have a total mass $10^{15}\,{\rm g} < M_{\rm dust} < 10^{17}\,{\rm
g}$ (see \S~\ref{subsec:dust}).  This circumstellar material could
originate either from the star or from an external source.  Plausible
external sources include an undetected old disk, comets, or a close-in
exoplanet.  This latter scenario would make CPVs extrasolar analogs of
the Jupiter-Io plasma torus (e.g.~Ref.~\cite{Bagenal1981}).

The other potential analog for the CPVs are the $\sigma$~Ori~E
variables, a rare subset of B stars with radiatively-driven winds
which accumulate into warped plasma tori
\cite{Townsend2005,Townsend2008}.  These tori tend to have dense
antipodal accumulations of plasma sculpted by tilted-dipole magnetic
fields, and the transits of these clumps produce broadband photometric
variability through bound-free scattering \cite{Townsend2005} and
Thomson scattering \cite{Berry2022}.  The plasma accumulates at
antipodes 180$^\circ$ apart because the deepest local minima in the
effective potential exist along the line of intersection between the
rotational and magnetic planes (see Equation~22 of
\cite{Townsend2005}).  For $\sigma$~Ori~E and almost all of its
analogs, the result is ``simple'' light curves that resemble those of
eclipsing binaries, and time-dynamic H$\alpha$ spectra similar to
those in Figure~\ref{fig:spec} \cite{Townsend2005,Townsend2008}.  The
two known exceptions, HD~37776 and HD~64740, show complex light curves
resembling CPVs \cite{Mikulasek2020,Bouma2024} and have
spectropolarimetric magnetic field maps indicating strong
contributions from higher-order magnetic moments
\cite{Kochukhov2011,Shultz2018}.  There are two implications: first,
the photometric complexity of CPVs may be a direct consequence of
magnetic fields with highly multipolar contributions.  Second, despite
this complexity, the observation (Figure~\ref{fig:spec}) of two
emission clumps separated 180$^\circ$ in phase is consistent with the
expected emission from a warped plasma torus.
